The ppmData R-package for setting up spatial point process models

### 2022-12-14 paper/paper.Rmd
updates to the manuscript

---
title: "The ppmData R-package for setting up spatial point process models"
tags:
  - Point Process
  - Quasi-random Sampling
  - Dirichlet Tesselation
  - Inhomogeneous Poisson Point Process
  - R package
authors:
  - name: Skipton N.C. Woolley
    orcid: 0000-0001-7022-0509
    affiliation: "1, 2"
  - name: Scott D. Foster
    orcid: 0000-0002-6719-8002
    affiliation: "3"
affiliations:
 - name: School of Ecosystems and Forestry Science, The University of Melbourne
   index: 1
 - name: Oceans & Atmosphere, CSIRO
   index: 2   
 - name: Data61, CSIRO
   index: 3    
citation_author: Woolley & Foster    
date: 07 July 2022
year: 2022
bibliography: paper.bib
header-includes: 
 - \usepackage{amssymb}
 - \newcommand{\R}{\mathbb{R}}
 - \newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
 - \usepackage{color}
 - \usepackage{bm}
output: rticles::joss_article
csl: apa.csl
journal: JOSS
editor_options: 
  chunk_output_type: console
---

@@

Spatial data, the locations of objects in a spatial region, is a common type of data recorded in ecology, environmental sciences and epidemiology amongst many others. An appropriate statistical approach for such data are spatial point process models [@cressie_statistics_1993; @baddeley_practical_2000]. Spatial point processes can be used to understand the location of objects to inform decision making and improve ecological, health and other socio-economic outcomes. For example, the locations of individuals from a species' population are typically not distributed completely at random, but rather are observed at locations with favorable habitat conditions for that species. We might expect these species sightings are distributed according to relevant covariates.

Inhomogeneous Poisson Process Models allow for the spatial intensity of a point process to vary across space, usually as a function of spatially referenced covariates. One challenge when fitting an Inhomogeneous Poisson Process Model is the integral used to described the intensity surface in the log-likelihood (as described below) has no known closed analytic solution. For Inhomogeneous Poisson Process Models, the Berman-Turner device [@berman_approximating_1992] is commonly used to approximate the model using quadrature within a weighted Poisson generalized linear model. The type of quadrature scheme can have important ramifications for overall model fitting and inference [@warton_poisson_2010]. Approximating the integral accurately often comes at an increased computation cost, where a large number of quadrature points are required to robustly estimate the integral [@renner_point_2015].

@@

Here we present the `ppmData` R package that is designed to setup a quadrature scheme using Dirichlet tessellation for fitting spatial point process models. Our initial intention was to develop multiple species (marked) quadrature in order to improve computation of multiple species Poisson point process models. Since then, the package has evolved into one which can be useful for fitting univariate response inhomogeneous spatial point process, as discussed in this paper. We demonstrate how to set up quadrature for inhomogeneous Poisson Process Models and provide a simple example of how we can fit an inhomogeneous Poisson Process Model using a `ppmData` object.

@@

There are a number of ways to setup a quadrature scheme to approximate the integral $\Lambda = \int_{\bm{s}\in\mathcal{A}}\lambda(\bm{s})d\bm{s}$. One common approach is to use a regular grid [@warton_poisson_2010], which creates a regular grid at a known resolution within a spatial window ($\mathcal{A} \subset R^2$). Despite the simplicity of the grid approach, it is not without detraction though. In particular, it is unlikely to handle edge effects well (when there is a consistent and potentially large pattern near the boundaries of $\mathcal{A}$). Then the grid will either be defined away from the edge, missing important trends, or defined on the edge, over-representing this trend. Also, another potential problem stems from the unlikely event that the spatial surface $\lambda(\bm{s}_i)$ has periodicity in its pattern aligned to the direction of the grid, such as repeated locations of mountain ridges. In that case, the grid-based quadrature will over- or under-estimate the integral depending on whether the grid is coincides with peaks or troughs of the surface.

@@

The package `spatstat` also allows a user to develop a quasi-random quadrature scheme, but this appears to be quite slow when used with Dirichlet weights for larger numbers of dummy (quadrature) points (\> 10000). For example, a regular window with 100000 quadrature points takes approximately 3.5 minutes to run in `spatstat` using `quadscheme` with `method=Dirichlet`. Our package computes the same Dirichlet tessellation in about 4.5 seconds. Typically, calculating a Dirichlet tessellation increases super-linearly and can become very slow for a large numbers of points. Part of the challenge when developing `ppmData` was to make this step efficient by a C++ implementation of a Delaunay triangulation radial sweep algorithm [@sinclair_s-hull_2016]. The Dirichlet tessellations are then calculated as the dual graph of the Delaunay triangulation. Areal weights for the quadrature points are calculated as the area of each Dirichlet tessellation polygon.

@@

In the `ppmData` package we have tried to make the generation of a quadrature scheme of a single or marked point process as easy as possible. We provide a simple interface for generating a Poisson point process data object and base all inputs and outputs using `terra` [@hijmans_terra_2021] and `sf` [@pebesma_simple_2018] packages. Users need to pass a set of coordinates from an observed point pattern, a window to define the point pattern region and a set of covariates observed across this region. The window is defined using a `terra` SpatRaster object, and it used to identify the point pattern region and where to generate the quadrature locations. Covariates are included as a multiple layered `terra` SpatRaster object that share the same extent and resolution as the window. The values of the covariates will be extracted at the locations of the point pattern and quadrature scheme. Whilst not unique to this package, the `ppmData` package can also generate grid and pseudo-random quadrature schemes as presented in Fig. 1.

@@

Here we demonstrate how to fit an Inhomogeneous Poisson Process Model using `glm` function and a `ppmData` object. This approach is very similar to the method presented in @warton_poisson_2010, except we are using a quasi-random quadrature. We use `glm` to fit an Inhomogeneous Poisson Process Model for simplicity, but if one wanted to fit an Inhomogeneous Poisson Process Model using different statistical machinery, then one could -- examples are: penalized regressions using `glmnet` [@friedman_regularization_2010] to achieve a `maxent` analysis [@renner_equivalence_2013], generalized additive models [ @wood_generalized_2017] or statistical learning approaches [@hastie_elements_2009].

@@

We present a snippet of example code for fitting an Inhomogeneous Poisson Process Model using `glm` from the R `stats` package [@r_core_team_r_2013]. We just need to extract the data.frame from the `ppmData` object and to specify a formula for the model. The only trick is making the response equal to `presence/weights` (the expectation). The right side of the formula will represent the covariates and their functional forms. In this example we specify independent 2$^{nd}$ degree polynomials for each covariate, except for the spatial coordinates X & Y which we include in the model as interacting 2$^{nd}$ degree polynomials. Once this is done, we define the weights in the `glm` function call and we now can fit an IPPM using `ppmData` and `glm`.

>>>>

---
title: "The ppmData R-package for setting up spatial point process models"
tags:
  - Point Process
  - Quasi-random Sampling
  - Dirichlet Tesselation
  - Inhomogeneous Poisson Point Process
  - R package
authors:
  - name: Skipton N.C. Woolley
    orcid: 0000-0001-7022-0509
    affiliation: "1, 2"
  - name: Scott D. Foster
    orcid: 0000-0002-6719-8002
    affiliation: "3"
affiliations:
 - name: Environment, CSIRO
   index: 1   
 - name: School of Ecosystems and Forestry Science, The University of Melbourne
   index: 2
 - name: Data61, CSIRO
   index: 3    
citation_author: Woolley & Foster    
date: 14 December 2022
year: 2022
bibliography: paper.bib
header-includes: 
 - \usepackage{amssymb}
 - \newcommand{\R}{\mathbb{R}}
 - \newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
 - \usepackage{color}
 - \usepackage{bm}
output: rticles::joss_article
csl: apa.csl
journal: JOSS
editor_options: 
  chunk_output_type: console
---

@@

Spatial data, the locations of objects in a spatial region, is a common type of data recorded in ecology, environmental sciences and epidemiology amongst many others. An appropriate statistical approach for data which can be described as points, such as plants or individuals of an animal population, are spatial point process models [@cressie_statistics_1993; @baddeley_practical_2000]. Spatial point processes can be used to understand the location of objects to inform decision making and improve ecological, health and other socio-economic outcomes. For example, the locations of individuals from a species' population are typically not distributed completely at random. Locations might be associated with favorable habitat conditions for the niche of that species. We might expect these species locations and the underlying spatial intensity are distributed according to relevant covariates, such as environmental conditions at those locations.

Inhomogeneous Poisson Process Models allow for the spatial intensity of a point process to vary across space, usually as a function of spatially referenced covariates. One challenge when fitting an Inhomogeneous Poisson Process Model is that the integral used to described the intensity surface in the log-likelihood (as described below) has no known closed analytic solution. For Inhomogeneous Poisson Process Models, the Berman-Turner device [@berman_approximating_1992] is commonly used to approximate the model using quadrature within a weighted Poisson generalized linear model. The type of quadrature scheme can have important ramifications for overall model fitting and inference [@warton_poisson_2010]. Approximating the integral accurately often comes at an increased computation cost, where a large number of quadrature points are required to robustly estimate the integral [@renner_point_2015].

@@

Here we present the `ppmData` R package that is designed to setup a quadrature scheme using Dirichlet tessellation for fitting spatial point process models. `ppmData` can setup a quadrature scheme of point process (e.g the locations of species) or a marked point process (e.g the locations of multiple species, where each location is associated with a specific species). In this paper, we demonstrate how to set up quadrature for inhomogeneous Poisson Process Models and provide a simple example of how we can fit an inhomogeneous Poisson Process Model using a `ppmData` object.

@@

There are a number of ways to setup a quadrature scheme to approximate the integral $\Lambda = \int_{\bm{s}\in\mathcal{A}}\lambda(\bm{s})d\bm{s}$. One common approach is to use a regular grid [@warton_poisson_2010], which creates a regular grid at a known resolution within a spatial window ($\mathcal{A} \subset R^2$). Despite the simplicity of the grid approach, it is not without detraction though. In particular, it is unlikely to handle edge effects well (when there is a consistent and potentially large pattern near the boundaries of $\mathcal{A}$). Then the grid will either be defined away from the edge, missing important trends, or defined on the edge, over-representing this trend. Also, another potential problem stems from the unlikely event that the spatial surface $\lambda(\bm{s}_i)$ has periodicity in its pattern aligned to the direction of the grid, such as repeated locations of mountain ridges. In that case, the grid-based quadrature will over- or under-estimate the integral depending on whether the grid coincides with peaks or troughs of the surface.

@@

The package `spatstat` also allows a user to develop a quasi-random quadrature scheme, but this appears to be quite slow when used with Dirichlet weights for larger numbers of dummy (quadrature) points (\> 10000). For example, a regular window with 100000 quadrature points takes approximately 3.5 minutes to run in `spatstat` using `quadscheme` with `method=Dirichlet`. Our package computes the same Dirichlet tessellation in about 4.5 seconds (Comparisons were from a Linux operating system using a Intel i7 8 core processor, with 16 GB of RAM). Typically, calculating a Dirichlet tessellation increases super-linearly and can become very slow for a large numbers of points. Part of the challenge when developing `ppmData` was to make this step efficient by a C++ implementation of a Delaunay triangulation radial sweep algorithm [@sinclair_s-hull_2016]. The Dirichlet tessellations are then calculated as the dual graph of the Delaunay triangulation. Areal weights for the quadrature points are calculated as the area of each Dirichlet tessellation polygon.

@@

In the `ppmData` package we have tried to make the generation of a quadrature scheme of a single or marked point process as easy as possible. We provide a simple interface for generating a Poisson point process data object and base all inputs and outputs using `terra` [@hijmans_terra_2021] and `sf` [@pebesma_simple_2018] packages. Users need to pass a set of coordinates from an observed point pattern, a window to define the point pattern region and a set of covariates observed across this region. The window is defined using a `terra` SpatRaster object, and it is used to identify the point pattern region and where to generate the quadrature locations. Covariates are included as a multiple layered `terra` SpatRaster object that share the same extent and resolution as the window. The values of the covariates will be extracted at the locations of the point pattern and quadrature scheme. Whilst not unique to this package, the `ppmData` package can also generate grid and pseudo-random quadrature schemes as presented in Fig. 1.

@@

Here we demonstrate how to fit an Inhomogeneous Poisson Process Model using the `glm` function and a `ppmData` object. This approach is very similar to the method presented in @warton_poisson_2010, except we are using a quasi-random quadrature. We use `glm` to fit an Inhomogeneous Poisson Process Model for simplicity, but if one wanted to fit an Inhomogeneous Poisson Process Model using different statistical machinery, then one could -- examples are: penalized regressions using `glmnet` [@friedman_regularization_2010] to achieve a `maxent` analysis [@renner_equivalence_2013], generalized additive models [ @wood_generalized_2017] or statistical learning approaches [@hastie_elements_2009].

@@

We present a snippet of example code for fitting an Inhomogeneous Poisson Process Model using `glm` from the R `stats` package [@r_core_team_r_2013]. We just need to extract the data.frame from the `ppmData` object and to specify a formula for the model. The only extra step is making the formula response equal to `presence/weights` (which gives us $z$ in the @berman_approximating_1992 approximation). The right side of the formula will represent the covariates and their functional forms. In this example we specify independent 2$^{nd}$ degree polynomials for each covariate, except for the spatial coordinates X & Y which we include in the model as interacting 2$^{nd}$ degree polynomials. Once this is done, we define the weights in the `glm` function call and we now can fit an IPPM using `ppmData` and `glm`.
